Covering Your Basis: Model Agnostic Probes of the Binary Black Hole Population

### 2022-09-28 src/tex/intro.tex
add landry/read NS citation

Observations of gravitational waves (GWs) from compact binary coalescences (CBCs) are becoming a regular occurrence, 
recently producing catalogs of 90 such detections \citep{GWTC1,gwtc2,GWTC3}. With this growing catalog size, we are steadily approaching a more complete  
understanding of the population properties of CBCs, just 7 years after the first observation \citep{o1o2_pop,o3a_pop,o3b_astro_dist}. 
In the coming fourth observing run of the LIGO-Virgo-KAGRA (LVK) collaboration detectors \citep{aLIGO,aVIRGO}, with the increased sensitivities, 
there could be a detection as often as every day \citep{LVK_prospects}. This large influx of incoming data will drastically increase what we can 
learn by combining all observations in order to study the compact object population. The history of these dense objects is imprinted on the properties 
of the population  of CBCs, allowing us to learn about their formation and evolution by the last few seconds of gravitational radiation emitted \citep{Zevin_2017}. 
The population distributions of merging compact objects inferred from gravitational-wave observations have been shown to be particularly useful in order to
measure cosmological parameters \citep{Farr_2019HUB,gwtc3_cosmo}, constrain modified gravitational wave propagation \citep{ModGWProp,ModGWProp2}, 
constrain a running Planck mass \citep{Lagos_runningPlanckMass}, look for evidence of ultralight bosons through superradiance \citep{Ng_Boson2021,GWTC2_superradiance_Ng}, 
constrain stellar nuclear reaction rates \citep{Farmer_2019,Farmer_2020}, look for primordial black holes \citep{Ng_2021,KenNgPBH2022}, 
and constrain the neutron star equation of state \citep{Golomb_EOS}. Through a better understanding of the mass, spin, and redshift distributions of 
compact binaries that will come with the increased catalog size, one can probe a wide range of different physical phenomena with even greater fidelity.

@@

The spin distributions of BBHs are particularly interesting since the two main formation scenarios produce distinctly different spin distributions \bruce{cite}. 
Isolated or field formation scenarios predict component spins that are preferentially aligned with the binaries orbital angular momentum, although some small 
misalignment can occur depending on the nature of the supernova kicks as each star collapses to a compact object \bruce{cite}. Dynamical formation scenarios may 
occur in dense environments where many-body interactions between a large number compact objects occur during the binary formation hardening the binary so that it 
will merge within a Hubble time. Thus, the dynamical channel produces an isotropic spin tilt distribution \bruce{cite}. BBH spins have also been of controversial 
interest recently, with different parametric approaches to modeling the spin distribution coming to different conclusions. The studies have disagreed on the possible existence 
of a significant zero-spinning subpopulation, along with the presence of significant spin misalignment (i.e. $\cos{\theta_i} < 0.0$) \citep{o3b_astro_dist,RouletGWTC2Pop,BuildBetterSpinModels,GWTC3MonashSpin,Callister_NoEvidence}. 
Another study recently showed that the spin tilt distribution inference is sensitive to modeling choices and may not peak at perfectly aligned spins, as is often assumed \citep{spinitasyoulike}. 
Other approaches to building improved models for BBH spins again, as with the mass models, focus on iterating through different simple parametric descriptions with 
increasing complexities \citep{BuildBetterSpinModels,Callister_NoEvidence,spinitasyoulike}. The spin distributions offer a nice use case for flexible data-driven models, as 
we present in this manuscript, both to help alleviate the current disputes in the literature and because there has been little work on flexible model development for the 
spin distribution of BBHs beyond assuming aligned spins thus far \citep{Tiwari_2021_a,Tiwari_2021_b}.

>>>>

Observations of gravitational waves (GWs) from compact binary coalescences (CBCs) are becoming a regular occurrence, 
recently producing catalogs of 90 such detections \citep{GWTC1,gwtc2,GWTC3}. With this growing catalog size, we are steadily approaching a more complete  
understanding of the population properties of CBCs, just 7 years after the first observation \citep{o1o2_pop,o3a_pop,o3b_astro_dist}. 
In the coming fourth observing run of the LIGO-Virgo-KAGRA (LVK) collaboration detectors \citep{aLIGO,aVIRGO}, with the increased sensitivities, 
there could be a detection as often as every day \citep{LVK_prospects}. This large influx of incoming data will drastically increase what we can 
learn by combining all observations in order to study the compact object population. The history of these dense objects is imprinted on the properties 
of the population  of CBCs, allowing us to learn about their formation and evolution by the last few seconds of gravitational radiation emitted \citep{Zevin_2017}. 
The population distributions of merging compact objects inferred from gravitational-wave observations have been shown to be particularly useful in order to
measure cosmological parameters \citep{Farr_2019HUB,gwtc3_cosmo}, constrain modified gravitational wave propagation \citep{ModGWProp,ModGWProp2}, 
constrain a running Planck mass \citep{Lagos_runningPlanckMass}, look for evidence of ultralight bosons through superradiance \citep{Ng_Boson2021,GWTC2_superradiance_Ng}, 
constrain stellar nuclear reaction rates \citep{Farmer_2019,Farmer_2020}, look for primordial black holes \citep{Ng_2021,KenNgPBH2022}, 
and to constrain physics of neutron stars \citep{Golomb_EOS,LandryRead_NS_Masses2021}. Through a better understanding of the mass, spin, and redshift distributions of 
compact binaries that will come with the increased catalog size, one can probe a wide range of different physical phenomena with even greater fidelity.

@@

The spin distributions of BBHs are particularly interesting since the two main formation scenarios produce distinctly different spin distributions \bruce{cite}. 
Isolated or field formation scenarios predict component spins that are preferentially aligned with the binaries orbital angular momentum, although some small 
misalignment can occur depending on the nature of the supernova kicks as each star collapses to a compact object \bruce{cite}. Dynamical formation scenarios may 
occur in dense environments where many-body interactions between a large number compact objects occur during the binary formation hardening the binary so that it 
will merge within a Hubble time. Thus, the dynamical channel produces an isotropic spin tilt distribution \bruce{cite}. BBH spins have also been of controversial 
interest recently, with different parametric approaches to modeling the spin distribution coming to different conclusions. The studies have disagreed on the possible existence 
of a significant zero-spinning subpopulation, along with the presence of significant spin misalignment (i.e. $\cos{\theta_i} < 0.0$) \citep{o3b_astro_dist,RouletGWTC2Pop,BuildBetterSpinModels,GWTC3MonashSpin,Callister_NoEvidence}. 
Another study recently showed that the spin tilt distribution inference is sensitive to modeling choices and may not peak at perfectly aligned spins, as is often assumed \citep{spinitasyoulike}. 
Other approaches to building improved models for BBH spins again, as with the mass models, focus on iterating through different simple parametric descriptions with 
increasing complexities \citep{BuildBetterSpinModels,Callister_NoEvidence,spinitasyoulike}. The spin distributions offer a nice use case for flexible data-driven models, as 
we present in this manuscript, both to help alleviate the current disputes in the literature and because there has been little work on flexible model development for the 
spin distribution of BBHs beyond the aligned spin dimension thus far \citep{Tiwari_2021_a,Tiwari_2021_b}.
